Two applications of logic programming to Coq

### 2020-11-04 papers/types2020/conc.tex
first reasonable draft of sec 3

% \subsection{Future work}
% \label{ssec:fut}
\todo{finish future work}
There are many avenues of developments for thisline of research.
The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
so far that we omitted a rule performing weak-head reduction (allowing
to handle directly existential goals) and more importantly that used
the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
use of constraints for data generation is well
developed~\cite{FioravantiPS15} and we could try to leverage to
improve PBT using partially instantiated terms: for example generating
say lists with at least two elements as \texttt{[1,2|Tail]}, without
recurring to needed narrowing as in in
LazySmallCheck~\cite{smallcheck}.

Extending our tactic to handle computations (\texttt{simp}) and/or to
return the partial proof state, unsolved goals in particular. Mention
\emph{Harpoon}, Beluga's new tactic language?

@@

\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is provate and full of stuff}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  Elpi

>>>>

% \subsection{Future work}
% \label{ssec:fut}
\todo{finish future work}
There are many avenues of developments for this line of research.
The code reported in Fig.~\ref{fig:augmented} is somewhat idealized in
so far that we omitted a rule performing weak-head reduction (allowing
to handle directly existential goals) and more importantly that used
the \emph{delay} mechanism. This a distinguished feature of Elpi, whose
usefulness has been advocated in ~\cite{DunchevCT16,coen19mscs}.  The
use of constraints for data generation is well
developed~\cite{FioravantiPS15} and we could try to leverage to
improve PBT using partially instantiated terms: for example generating
say lists with at least two elements as \texttt{[1,2|Tail]}, without
recurring to needed narrowing as in in
LazySmallCheck~\cite{smallcheck}.

Extending our tactic to handle computations (\texttt{simpl}) and/or to
return the partial proof state, unsolved goals in particular. Mention
\emph{Harpoon}, Beluga's new tactic language?

@@

\todo{Link to the repo. Which one? \url{https://github.com/proofcert/fpc-elpi} is private and full of stuff}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  Elpi simp LJF

### 2020-11-04 papers/types2020/sec3.tex
first reasonable draft of sec 3

Figure~\ref{fig:augmented} contains an implicitely focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgements for the two categories. In our case, where proof search steps are
viewed as computation, the term judgement corresponds to a goal-reduction
step while the list judgement corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
don't need to have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments, and as is usual in Pure Type
Systems we have the same grammar for terms and types:
%, as shown in figure~\ref{fig:terms}.
\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
%\begin{figure}
\begin{align*}
t,u ::= & k\ l \;|\; \forall x : t.\, u \\
l ::= & [] \;|\; (t\ ::\ l)
\end{align*}
% \caption{Grammar of terms}
% \label{fig:terms}
% \end{figure}

\todo{This does not see quite right: D vs A, consider say even}
The proof system is parametrized by  external set of inductive definitions available in
the (implict) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

%
%
%done Explain relations with other term assignments. Note that we do not use
% explicit substitutions (we are cut free)
% Explain the choice of not having certificates represent proof terms.
\todo{  Explain that the unfold clause handles
defined atoms and defined connectives (conjunction, disjunction, existential
and also equality): this may be less efficient and proof-theoretically induce
pointless phase changes, but it is arguably very elegant e simplifies proof
terms that arand inherited by Coq }

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \lsti{prod ``x'' A (x\ B x)}; when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implict} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Now safe-dest-app is hiiden inside \texttt{definition}. Put the rest of the code in the appendix?}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

@@

\begin{metanote}
  AM stops here
\end{metanote}
\subsection{A prolog-like tactic}

\begin{metanote}
  Let's dump it here and see how it looks
\end{metanote}

\begin{lstlisting}
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _Ctx Ev G _Who] _OutGoals :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
}}.
\end{lstlisting}
The following example show how we can do FPC-driven logic programming in Coq and return
a Coq term
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter capability of solving
goals by \lsti{reflexivity} and the modularity capability of using
\lsti{Hints}, but it is not compelled to use depth-first search, since it follows the dictate of the given FPC. For example the FPC can provide a \emph{trace}, which may be more customable that the one offered by the \lsti{Debug} option.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity

>>>>

Figure~\ref{fig:augmented} contains an implicitly focused proof system  for a small fragment of the Calculus of Inductive Constructions that corresponds to Horn logic programming over pure \lsti{Inductive} definitions.
%
It is a sequent calculus inspired by the calculus for
proof search in Pure Type Systems introduced in \cite{LengrandDM06}, based in
 turn on ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similarly to that calculus, we have a term
language that includes terms and lists of terms, and two distinct typing
judgments for the two categories. In our case, where proof search steps are
viewed as computation, the term judgment corresponds to a goal-reduction
step while the list judgment corresponds to a backchaining step. This coincides
with the idea behind the \emph{spine calculus}~\cite{Cervesato97tr}.

It may be at first surprising that there are no introduction rules for
propositional connectives, nor equality for that matter. However, one of the beauty of the Calculus of Inductive Construction is that they are in fact defined inductively and therefore the \emph{unfold} rule will handle those; this also simplifies proof-terms.

Since we are only dealing with types that correspond to Horn clauses, and
atomic types are inductively defined, we do not need a $\forall$ rule on the right, although the proof theory would gladly allow it. This means, at the term level, that we
do not have variables in our grammar of terms. Terms are always
applied to a (possibly empty) list of arguments, and as is usual in Pure Type
Systems we have the same grammar for terms and types:
%, as shown in figure~\ref{fig:terms}.
\todo{Let's think about that: we need A at least, but A is also a metavariable for atoms}
%\begin{figure}
\begin{align*}
t,u ::= & k\ l \;|\; \forall x : t.\, u \\
l ::= & [] \;|\; (t\ ::\ l)
\end{align*}
% \caption{Grammar of terms}
% \label{fig:terms}
% \end{figure}

\todo{This does not see quite right: D vs A, consider say even}
The proof system is parameterized by  external set of inductive definitions available in
the (implicit) global context: we denote by $\Pscr_A$ the set of type declarations for the
constructors of type $A$, of the form $k : D$ where $D$ is of the form
$\forall x_1 : t_1.\, \dots \forall x_n : t_n.\, D\ u_1 \dots u_m$. We denote
by $\mathcal{A}$ the set of arity declaration for the inductive types.

%
%
%done Explain relations with other term assignments. Note that we do not use
% explicit substitutions (we are cut free)
% Explain the choice of not having certificates represent proof terms.
% \todo{  Explain that the unfold clause handles
% defined atoms and defined connectives (conjunction, disjunction, existential
% and also equality): this may be less efficient and proof-theoretically induce
% pointless phase changes, but it is arguably very elegant e simplifies proof
% terms that are inherited by Coq }

@@

Note that \lsti{prod} encodes dependent products taking a name for pretty printing,  a term and a \lP function from
terms to terms, namely $\forall x : A, B$ is encoded by \texttt{prod ``x'' A (x\ B x)};
when, in the implementation of the product-left rule, we
apply $B$ to the variable \lsti{Tm},
we get a term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In this sense, our calculus uses \emph{implicit} substitutions, rather than explicit ones as in LJT and PTSC's tradition, the more since proof search in our application is cut-free. 
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
\todo{Now safe-dest-app is hidden inside \texttt{definition}. Put the rest of the code in the appendix?}
The unfolding rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} in order to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} in order to access the global
environment of inductive definitions and query for information about them.

@@

\subsection{A prolog-like tactic}

\begin{metanote}
Keep this? Perhaps explain Ev = Term?
\end{metanote}

Thanks to the Coq-Elpi interface, in particular to the ``main''
proocedure \lsti{solve}, we can package the \lP's code for the checker
as a tactic that can be called as any other tactic in a Coq script.

\begin{lstlisting}
Elpi Tactic dprolog.
Elpi Accumulate lp:{{
  solve [str ''height'', int N] [goal _  Ev G _] _ :-
    coq.say "Goal:" {coq.term->string G},
    check (qheight N) (go G Term),
    Ev = Term,
    coq.say "Proof:" {coq.term->string Ev}.
...
/* Other clauses for different fpc omitted */
}}.
\end{lstlisting}
The following example show how we can use the above tactic to do
FPC-driven logic programming in Coq and return a Coq proof-term:
\begin{lstlisting}
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : is_nat x -> insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 [0;1] R.
eexists.
elpi  dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 [0; 1] R) [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (i_c 2 1 [] [2] (i_n 2)) : exists R : list nat, insert 2 [0; 1] R
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable version of
Coq's \lsti{auto}: it currently lacks the latter's capability of
solving goals by \lsti{reflexivity} as well as the modularity of using
\lsti{Hints} as introduction rules; on the bright side, it is not
restricted to depth-first search, since it follows the dictate of the
given FPC. For example the FPC can provide a \emph{trace} that may be
more customable that the one offered by the \lsti{auto}'s hard-wired
\lsti{Debug} facility.
%
%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:

% LocalWords:  backchaining disjunct LJT dep qheight qsize orE someE
% LocalWords:  Elpi modularity Herbelin bc PTSC's dprolog
